Interrogating the linear variable importance of local explanations of black-box models with animated linear projections

### 2021-12-04 _paper.tex
images to .png; to allow for printing .pdf (.pdf format with alpha causes `flattening` to hang).
Edits before restructuing to classification/regression case.

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of explainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an R package \textbf{cheem} available at \href{https://github.com/nspyrison/cheem}{github.com/nspyrison/cheem}.
\end{abstract}

@@

Breiman and Shmueli Shmueli (2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling is performed to predict new or future out-of-sample observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, the interpretable is a key feature for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use of a model has important implications for which methods are used and the development of those models.

Black-box models are becoming increasingly common, but not without their share of controversy and issues Kodiyan (2019). Applications have been known to reflect common biases against sex Duffy (2019), race (Larson et al. 2016), and age (Díaz et al. 2018). These are all-too common issue, when endogenous or exogenous biases in the training data and identified and mimicked by models. Another issue is that of data-drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions Salzberg (2014). Such issues highlight the need to make models fair, accountable, ethical, and transparent which has led to the movement of XAI Arrieta et al. (2020).

@@

To illustrate our work we apply the model-agnostic explanation SHAP Štrumbelj and Kononenko (2014). The exact details of SHAP are tangent to the ideas of this work, but suffice it to say that SHAP approximates variable importance by taking the median importance over permutations of the explanatory variables. To be exact we apply a variant that enjoys a lower computational complexity, known as tree SHAP (S. M. Lundberg, Erion, and Lee 2018).

In multivariate data visualization a \emph{tour} S. Lee et al. (2021) is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to a projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of the structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

There are various types of tours, which are distinguished by the sequence of projection bases. In a \emph{manual} tour Spyrison and Cook (2020) this path is defined by changing the contribution of a selected variable. Applying tours in conjunction with models has been previously done, \emph{ie} for exploring various statistical model fits (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path Silva, Cook, and Lee (2021).

@@

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{Ceteris paribus}(Biecek 2020). Latin for ``other things held constant'' or ``all else unchanged,'' is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation a of the factual observed case by varying the basis, the angle of the data object. Another, but related difference is that this geometric approach maintains orthogonal dimensions. That is to say when the contribution of some variables decrease the the contributions of others necessarily increases.

The rest of this paper is organized as follows. The next section \protect\hyperlink{sec:explanations}{Local explanation statistics} covers the background of the local explanation SHAP and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into the animation and what it shows. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis and discusses the backend details of the package and preprocessing. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{discussion} of the insights we draw from classification and regression tasks.

@@

Consider a highly non-linear model. At face value its hard to say which variable are sensitive to the crossing of classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance at in vicinity of one observation.

Figure 6 of Arrieta et al. (2020) gives a broad summarization of the taxonomy and literature of explanation techniques. This includes a large number of model-specific explanations such as deepLIFT, Shrikumar, Greenside, and Kundaje (2017) a popular recursive method for estimating importance in neural networks. There are a fewer number of model-agnostic explanations, of which LIME, (Ribeiro, Singh, and Guestrin 2016) SHAP, (S. Lundberg and Lee 2017) and their variants are popular.

These local explanation are used in a variety of ways depending on the data. For images saliency maps (Simonyan, Vedaldi, and Zisserman 2014) a heatmap can be used to show which pixels were important for distinguishing pictures of wolfs or huskies for instance. In text analysis saliency can be used to highlight influential words (Vanni et al. 2018). In the case of numeric regression they can be used to explain variable additive contributions from intercept to prediction (Ribeiro, Singh, and Guestrin 2016).

@@

SHaply Additive exPlanations, or SHAP approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory where Shapley devised a method to evaluate individual's contribution to cooperative games by permuting the players that contribute to the score (Shapley 1953).

To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season (Leone 2020). We have 5000 observations of 9 skill measures (after aggregating variable with high correlation). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players recieve a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions as well, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry with respect to the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

{\centering \includegraphics{./figures/shap_distr_bd}

@@

\caption{Illustration of the distribution of SHAP attributions, the SHAP values, and a breakdown plot, the typical visual of SHAP local explanations. For FIFA 2020 data, of a random forest model regressing wages from 9 skill attributes for a star offensive and defensive player. a) The players have very different wages. b) Shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The median of these distributions are the final SHAP values, notice that that the variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the position of the player. c) Breakdown plots of the observations the explanation used to additively explain the difference between the intercept and prediction}\label{fig:shapdistrbd}
\end{figure}

@@

It is insightful to explore these two approximations against a visual of the model; prediction by observation is added. Linked brushing and preserved aesthetic features such as circling misclassifed observations helps link information from the different spaces together.

@@

The primary observation identified via the global view is foundational to the production of the cheem tour. Namely, the linear attribution of that variable is used as a 1D projection basis. This is the approximate contributions of the variables that this model uses to justify its prediction for the observation.

The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \ref{fig:cheemtourclass} we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

In the regression case we compare an observation with a particularly extreme residual as compared with a neighboring point that is more accurately predicted. The global tour is primarily the same without the misclassified circles, but the tooltip now shows the observation's residual. We simulate data, 5 variable from the uniform distribution (between 0 and 5) and \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). We fit a random forest and extract the all tree SHAP values. The first thing we notice from the global view is the shape difference between the data- and attribution-spaces, which are primarily circular are triangular respectively. We also see that the random forest fit the data quite well viewing the predictions across the observations. Though linked brushing we find that the attribution space is highly correlated with the \(y\) value. The highest values occupy 1 corner of the triangle, the lowest values cluster on the opposite side and the middle values form a `)' shape filling out the triangle in attribution space. In the regression case the tour view is a bit different, this is illustrated in figure \ref{cheemtourreg}. The the current frames basis and distribution are shown to the left. The horizontal positions of the two scatter plots are the resulting 1D projection of the basis while the heights are fixed to the observed y and the residuals.

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable, an optional aesthetic variable (color/shape)  can be mapped typically a categorical variable.
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} (Liaw and Wiener 2002) to mitigate the runtime of our local explanation which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP applicable to tree-based models. This is done with the package \textbf{treeshap} (Kominsarczyk et al. 2021), hosted on GitHub only]. The global view shows all observations in attribution space requiring that we must extract the variable weightings from \emph{all} observations rather than just one.
    \item \textbf{Global view:} The data- and attribution-spaces are approximated as their the first two principal components.
\end{itemize}

The time to preprocess the data will vary significantly with the choice of model and local explanation. However, for reference, the FIFA data, 5000 observations of 9 explanatory variables, took 0.6 seconds to create PCA for both the data and attribution spaces. On the same data, with modest hyper parameters a random forest model fit in 2.9 seconds while extracting the tree SHAP values of each observation took 254 seconds combined. These runtimes were from a non-parallelized R session on a modern laptop, but suffice it to say that the bulk of the time will be spent on the local attribution. This makes tree SHAP a good candidate to start with. The package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes that are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} \emph{TODO:XXX site github? cran?}. Preprocessing was facilitated with models created via \textbf{randomForest} {[}liaw\_classification\_2002{]}, and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021).
The application is made with \textbf{shiny} (Chang et al. 2021). The tour visual is an extension of \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive HTML widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and
the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) was a huge boon to understanding local explanations and how to apply them.

@@

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Download the package:}
\FunctionTok{install.packages}\NormalTok{(}\StringTok{"cheem"}\NormalTok{, }\AttributeTok{dependencies =} \ConstantTok{TRUE}\NormalTok{)}
\DocumentationTok{\#\# Try the app yourself:}
\NormalTok{cheem}\SpecialCharTok{::}\FunctionTok{run\_app}\NormalTok{()}
\DocumentationTok{\#\# Procss your own data for the app by following the examples in:}
\NormalTok{?cheem}\SpecialCharTok{::}\NormalTok{cheem\_ls}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{penguins-species-classification}{%
\subsection{1) penguins, species classification}\label{penguins-species-classification}}

Palmer penguins data Horst, Hill, and Gorman (2020) consist of 330 observations across four physical measurements of 3 species of penguins foraging near Palmer Station, Antarctica. A random forest model was fit, classifying the species of the penguin given the physical measurements.

@@

The chocolates dataset consists of 88 observations of 10 nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. With this data, we can see if a manufacturer is giving an accurate portrayal of the chocolate. We are curious to see if there are chocolates that nutritionally look like milk chocolates that are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We'll select chocolate labeled dark, through predicted to be milk chocolate with a comparison with a chocolate labeled 85\% cocoa.

@@

The 2020 season FIFABiecek (2018), contains many skill measurements of soccer/football players along with wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit with 5000 observations of the 9 skill aggregates before being thinned to 500 players to gate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a leading defensive fielder (V. van Dijk), the same observations used in figure @figure(fig:shapdistrbd).

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader MI\^{}2 lab group for the \textbf{DALEX} ecosystem of \textbf{R} and \textbf{Pyhton} packages. This research was supported in part by Australian government Research Training Program (RTP) scholarships.

The namesake, Cheem, refers to a fictional race of humanoid trees in Doctor Who lore. Given that \textbf{DALEX} pulls on from that universe and we originally apply tree SHAP explanations (specific to tree-based models) we found it fitting.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an \textbf{R} package \textbf{cheem} available at \href{https://github.com/nspyrison/cheem}{github.com/nspyrison/cheem}.
\end{abstract}

@@

Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling is performed to predict new or future out-of-sample observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, the interpretable is a key feature for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use of a model has important implications for which methods are used and the development of those models.

Black-box models are becoming increasingly common, but not without their share of controversy and issues (O'neil 2016; Kodiyan 2019). Applications have been known to reflect common biases against sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018). Such issues occur when biases existent in the training data, the model picks up on this influence on the response variable which is then built into the model. Another issue is that of data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions (Lazer et al. 2014; Salzberg 2014). Such issues highlight the need to make models fair, accountable, ethical, and transparent which has led to the movement of XAI (Adadi and Berrada 2018; Arrieta et al. 2020).

@@

To illustrate our work we apply the model-agnostic explanation SHAP (Strumbelj and Kononenko 2010; Štrumbelj and Kononenko 2014). The exact details of SHAP are tangent to the ideas of this work, but suffice it to say that SHAP approximates variable importance by taking the median importance over permutations of the explanatory variables. To be exact we apply a variant that enjoys a lower computational complexity, known as tree SHAP (S. M. Lundberg, Erion, and Lee 2018).

In multivariate data visualization a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of the structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

There are various types of tours, which are distinguished by the sequence of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020) this path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

@@

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{Ceteris paribus} (Biecek 2020). Latin for ``other things held constant'' or ``all else unchanged,'' is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual observed case by varying the basis, the angle of the data object. Another difference is that the basis is restricted on orthonormality. That is to say when the contribution of one variable decreases the contributions of others necessarily increases such that there is a full component in that direction.

The rest of this paper is organized as follows. The next section \protect\hyperlink{sec:explanations}{Local explanation statistics} covers the background of the local explanation SHAP and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into the animation and what it shows. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis, and discusses the backend details of the package and preprocessing. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{Discussion} of the insights we draw from classification and regression tasks.

@@

Consider a highly non-linear model. At face value it is hard to say which variable(s) are sensitive to crossing of classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

Figure 6 of Arrieta et al. (2020) gives a broad summarization of the taxonomy and literature of explanation techniques. This includes a large number of model-specific explanations such as deepLIFT, (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017) a popular recursive method for estimating importance in neural networks. There are a fewer number of model-agnostic explanations, of which LIME, (Ribeiro, Singh, and Guestrin 2016) SHAP, (S. Lundberg and Lee 2017) and their variants are popular.

These instance level explanations are used in a variety of ways depending on the data. In images saliency maps overlay or offset a heatmap indicating which pixels were important (Simonyan, Vedaldi, and Zisserman 2014). For instance, or the presence of snow may be highlighted when distinguishing if a picture of contains a wolf or husky. In text analysis, Word-level contextual sentiment analysis can be used to highlight sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they can be used to explain variable additive contributions from intercept to prediction (Ribeiro, Singh, and Guestrin 2016).

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory where Shapley devised a method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score (Shapley 1953).

To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season (Leone 2020). We have 5000 observations of nine skill measures (after aggregating variable with high correlation). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions as well, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry with respect to the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

{\centering \includegraphics[width=29.17in]{./figures/shap_distr_bd}

@@

\caption{Illustration of the distribution of SHAP attributions and a breakdown plot. From FIFA 2020 data, a random forest model regresses wages from 9 skill attributes for a star offensive and defensive player. a) The players have very different wages. b) Shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The median of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the position of the player. c) Breakdown plots of the observations the explanation used to additively explain the difference between the intercept and prediction}\label{fig:shapdistrbd}
\end{figure}

@@

It is insightful to explore these two approximations against a visual of the model; prediction by observation is added. Linked brushing and preserved aesthetic features such as circling misclassified observations help link information from the different spaces together.

@@

The primary observation identified via the global view is foundational to the production of the cheem tour. Namely, the linear attribution of that variable is used as a 1D projection basis. This is the approximate contribution of the variables that this model uses to justify its prediction for the observation.

The SHAP values of the full dataset are shown as horizontal parallel coordinate plots at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of a bar is mapped to the variable's contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \ref{fig:cheemtourclass} we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

In the regression case, we compare an observation with a particularly extreme residual as compared with a neighboring point that is more accurately predicted. The global tour is primarily the same without the misclassified circles, but the tooltip now shows the observation's residual. We simulate data, five variables from the uniform distribution in the support {[}0, 5{]} and \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). We fit a random forest and extract the tree SHAP values from all observations. The first thing we notice from the global view is the shape difference between the data- and attribution-spaces, which are primarily circular are triangular respectively. We also see that the random forest fit the data quite well viewing the predictions across the observations. Though linked brushing we find that the attribution space is highly correlated with the \(y\) value. The highest values occupy one corner of the triangle, the lowest values cluster on the opposite side, and the middle values form a `)' shape filling out the triangle in attribution space. In the regression case, the tour view is a bit different, this is illustrated in figure \ref{fig:cheemtourreg}. The current frames basis and distribution are shown to the left. The horizontal positions of the two scatter plots are the resulting 1D projection of the basis while the heights are fixed to the observed y and the residuals.

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable, an optional aesthetic variable (color/shape)  can be mapped typically a categorical variable.
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} (Liaw and Wiener 2002) to mitigate the runtime of our local explanation which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP applicable to tree-based models. This is done with the package \textbf{treeshap} (Kominsarczyk et al. 2021), hosted on GitHub only]. The global view shows all observations in attribution space requiring that we must extract the variable weightings from \emph{all} observations rather than just one.
    <!-- \item \textbf{Global view:} The data- and attribution-spaces are approximated as their the first two principal components. -->
\end{itemize}

The time to preprocess the data will vary significantly with the choice of model and local explanation. For reference, the FIFA data, 5000 observations of 9 explanatory variables, took 2.9 seconds to fit a random forest models of modest hyperparmeters. Extracting the tree SHAP values of each observation took 254 seconds combined. PCA and statistics of the variables and attribution took 0.6 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice it to say that the bulk of the time will be spent on the local attribution. This makes tree SHAP, with its reduced comupational complexity a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} \emph{TODO:XXX include cran URL after published}. Preprocessing was facilitated with models created via \textbf{randomForest} {[}liaw\_classification\_2002{]}, and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021). The application is made with \textbf{shiny} (Chang et al. 2021). The tour visual is an extension of \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive HTML widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) was a huge boon to understanding local explanations and how to apply them.

@@

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Download the package:}
\FunctionTok{install.packages}\NormalTok{(}\StringTok{"cheem"}\NormalTok{, }\AttributeTok{dependencies =} \ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{library}\NormalTok{(}\StringTok{"cheem"}\NormalTok{)}
\DocumentationTok{\#\# Try the app:}
\FunctionTok{run\_app}\NormalTok{()}
\DocumentationTok{\#\# Bring your own data: follow the examples in cheem\_ls()}
\NormalTok{?cheem\_ls}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{penguin-species-classification}{%
\subsection{1) Penguin, species classification}\label{penguin-species-classification}}

Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) consist of 330 observations across four physical measurements of three species of penguins foraging near Palmer Station, Antarctica. A random forest model was fit, classifying the species of the penguin given the physical measurements.

@@

The chocolates dataset consists of 88 observations of 10 nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. With this data, we can see if a manufacturer is giving an accurate portrayal of the chocolate. We are curious to see if there are chocolates that nutritionally look like milk chocolates that are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We select a chocolate labeled dark, through predicted to be milk chocolate with a comparison with a chocolate labeled 85\% cocoa.

@@

The 2020 season FIFA (Leone 2020; Biecek 2018), contains many skill measurements of soccer/football players along with wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit with 5000 observations of the 9 skill aggregates before being thinned to 500 players to gate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a leading defensive fielder (V. van Dijk), the same observations used in figure \ref{fig:shapdistrbd}.

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader \(\text{MI}^\text2\) lab group for the \textbf{DALEX} ecosystem of \textbf{R} and \textbf{Pyhton} packages. This research was supported in part by Australian government Research Training Program (RTP) scholarships.

The namesake, Cheem, refers to a fictional race of humanoid trees in Doctor Who lore. \textbf{DALEX} pulls on from that universe and we originally apply tree SHAP explanations which are specific to tree-based models.
